---
title: "Applied Regression and Time Series Analysis: Lab 2"
author: "Chris Bennett"
date: "November 30, 2016"
output: pdf_document
---

###Instructions:

#### Forecast Inflation-Adjusted Gas Price
During 2013 amid high gas prices, the Associated Press (AP) published an article about the U.S. inflation-adjusted price of gasoline and U.S. oil production. The article claims that there is "*evidence of no statistical correlation*" between oil production and gas prices. The data was not made publicly available, but comparable data was created using data from the Energy Information Administration. The workspace and data frame *gasOil.Rdata* contains the U.S. oil production (in millions of barrels of oil) and the inflation-adjusted average gas prices (in dollars) over the date range the article indicates.

You have three tasks for this exericse, and both tasks need the use of the data set *gasOil.Rdata*.

##Task 1: Create a **SARIMA-type** model to forecast the inflation-adjusted gas prices, and use this model to forecast the inflation-adjusted gas price for the next two years.

```{r}
######################################################################
# Directory         : 
# Program Name      : Section1_Lab2_BennettC.Rmd
# Original Developer: Chris Bennett
# Last Updated by   : Chris Bennett
# Last Updated      : 11/30/2016
#####################################################################

#####################################################################
# Setup

# Set working directory
setwd("C:/data")
getwd()

# Set Numeric Value Display
# See reference from https://stat.ethz.ch/R-manual/R-devel/library/base/html/options.html
options(digits=7) # Set the printed number of digits to 7. Default is 7. 
#options("scipen" = 10)

# Set memory limit
memory.limit(50000000)

################################################################
# Load Libraries

library(astsa)    # Time series package by Shummway and Stoffer
library(zoo)      # time series package
library(forecast)
library(psych)
library(car)
library(lmtest)
library(sandwich)
library(ggplot2)
require(scales)
require(GGally)
require(tseries)
require(fGarch)   # Garch package
require(quantmod) # Financial time series package
library(vars)
#require(tseries)
#require(MASS)
#require(graphics)
#require(corrgram)

################################################################

#General Exploratory Data Analysis

#clear out any loaded objects
rm(list=ls())
ls()

#load data set
load("gasOil.Rdata")

#begin EDA
ls(gasOil)
str(gasOil)

#number of columns and rows
nrow(gasOil)
ncol(gasOil)

describe(gasOil)
summary(gasOil)
head(gasOil)
tail(gasOil)

#look for missing values in each column
sapply(gasOil, function(x) sum(is.na(x)))

price.ts = ts(gasOil[,3], start=c(1978,1,1), end=c(2012,2,1), freq=12)
production.ts = ts(gasOil[,2], start=c(1978,1,1), end=c(2012,2,1), freq=12)

#price.z = zoo(x=as.numeric(gasOil[,3]), order.by=as.Date(gasOil[,1]))

#evaluate the zoo data object
class(price.ts)
str(price.ts)
head(price.ts,10)

index(price.ts)
coredata(price.ts)
start(price.ts)
end(price.ts)

par(mfrow=c(2,2))
plot(price.ts, col="blue", main="Fig 1: Inflation-Adjusted Gas Prices Series Plot")
hist(price.ts, col="gray", main="Fig 2: Inflation-Adjusted Gas Prices Histogram")
acf(price.ts, main="Fig 3: ACF: Inflation-Adjusted Gas Prices")
pacf(price.ts, main="Fig 4: PACF: Inflation-Adjusted Gas Prices")

price.diff <- diff(price.ts, lag=12)

par(mfrow=c(2,2))
plot(price.ts, col="blue", main="Fig 1: Inflation-Adjusted Gas Prices Series Plot")
plot.ts(price.diff); title("Fig 2: First Difference of the Gas Prices Series")
acf(price.diff, main="Fig 3: ACF: First Difference of the Gas Prices Series")
pacf(price.diff, main="Fig 4: PACF: First Difference of the Gas Prices Series")

fit1 <- arima(price.ts, order=c(2,1,2)) 
fit1

pvar<-1:3
dvar<-1:2
qvar<-1:2

OrderGrid<-expand.grid(pvar,dvar,qvar)

ModFit <- function(x, dat){
          m=Arima(dat, order=c(x[[1]], x[[2]], x[[3]]))
          return(m)
} 

Fits <- plyr::alply(OrderGrid, 1, ModFit, dat = price.ts)

Fits

```

##Overall Observations

* There are 410 rows of data, and 3 columns
* The 3 columns are Date, Production, and Price
* There appear to be no missing values from the data frame

##Variables

* Dates: 
    * Appear to be monthly, specifically the first day of each month
    * First Date: January 1, 1978
    * Last Date: February 1, 2012
* Production:
    * Mean: 210.01 (standard deviation of 41.88)
    * Min: 119.41
    * Max: 283.25
* Price:
    * Mean: 2.39 (stdev of 0.7)
    * Min: 1.33
    * Max: 4.43

####Stationarity

* The series does not appear to be stationary in mean or variance.
* The histogram of values shows a non-normal distribution with the largest number of the values between 1.5 and 2 with a significant positivie skew.
* The plot of the series shows a large continuous negative shock between 1980 and 1987, a fairly consistent sequence between 1987 and 2005, then a large positive shock after 2005.  
* The fact that this series is not stationary is further supported by the Autocorrelation Function showing significant (albeit decreasing) correlation for all lags displayed.

####Seasonality

* There appears to be some seasonal variation in the general plot that we will explore further.
* The ACS does show some slight "waves" in the correlations between lags, suggesting possible seasonality.

##Task 2: Create a *multivariate* time series model that can be used to predict/forecast inflation-adjusted gas prices, and use this model to forecast the inflation-adjusted gas price for the next two years

```{r}
# Fitting a linear trend to the Gas Price Data
# Regress Price on time
summary(price.fit_lm <- lm(price.ts ~ time(price.ts)))
price.fit_lm$coeff
price.fit_lm$qr
price.fit_lm$fitted.values

# Fitting a linear trend to the Gas Production Data
# Regress Production on time
summary(production.fit_lm <- lm(production.ts ~ time(production.ts)))
production.fit_lm$coeff
production.fit_lm$qr
production.fit_lm$fitted.values

#Plot the regression lines for both Price and Production
par(mfrow=c(2,1))
plot(price.ts, col="blue", main="", xlab="Year", ylab="Gas Price")
title("Inflation-Adjusted Gas Prices")
abline(price.fit_lm) # add the fitted regression line to the plot

plot(production.ts, col="blue", main="", xlab="Year", ylab="Gas Production")
title("Inflation-Adjusted Gas Production")
abline(production.fit_lm) # add the fitted regression line to the plot

#Look at the first few records for both price and production
head(cbind(price.ts, production.ts))

#Evaluate Correlation between price and correlation
cor(price.ts, production.ts)

#Plot the price and production time series
par(mfrow=c(2,2))
plot.ts(price.ts); title("Time-Series Plot of Price")
plot.ts(production.ts); title("Time-Series Plot of Production")
ts.plot(ts(price.ts),ts(production.ts)); title("Time-Series Plot of Price and Production")
plot(price.ts,production.ts); title("Price vs Production")

# Show summary statistics, even though they might not be useful for a timeseries
summary(cbind(price.ts, production.ts))

#Plot the autocorrelation fuction and partial autocorrelation function for both price and production for 36 lags
par(mfrow=c(2,2))
acf(price.ts, main="", 36); title("ACF of Price")  
acf(production.ts, main="", 36); title("ACF of Production")
pacf(price.ts, main="", 36); title("PACF of Price")
pacf(production.ts, main="", 36); title("PACF of Production")

#More EDA on Price time series (plot, histogram, acf, and pacf)
par(mfrow=c(2,2))
plot.ts(price.ts)
hist(price.ts)
acf(price.ts)
pacf(price.ts)

#More EDA on Production time series (plot, histogram, acf, and pacf)
par(mfrow=c(2,2))
plot.ts(production.ts)
hist(production.ts)
acf(production.ts)
pacf(production.ts)

#Run the Augmented Dickey-Fuller Test to determine if Unit Root is present for price and production
adf.test(price.ts)
adf.test(production.ts)
#none of these tests fail to reject the null hypothesis that each of these series is unit root.

#Phillips-Ouliaris Cointegration Test
po.test(cbind(price.ts,production.ts))

#Aggregate into an annualized series
cbind(price.ts[1:12],production.ts[1:12])
cbind(aggregate(price.ts), aggregate(production.ts))
cbind(sum(price.ts[1:12]),sum(production.ts[1:12]))

tempx<-aggregate(price.ts); tempy<-aggregate(production.ts)
cbind(tempx[1],tempy[1]) # which is the same as sum(price.ts[1:12]) and sum(production.ts[1:12])

#Plot Aggregate annualized values for Price and Production
par(mfrow=c(1,1))
plot(as.vector(tempx), as.vector(tempy), xlab="Price", ylab="Production", col="blue")
title("Annual Price vs. Annual Production")

#remove objects
rm(tempx, tempy)

# Compute the correlation between the 2 series
cat("Corr(Price , Production): ", cor(price.ts,production.ts))

#Investigating the cointegrated model

# 1. Fit a linear regression model
GAS.lm <- lm(price.ts ~ production.ts)
summary(GAS.lm)
# The slop coefficient estimate is highly significant, but the overall model and the coefficient are not signficant. # The R-square is very low.

# 2. Obtain the residuals
GAS.res <- resid(GAS.lm)
summary(GAS.res)  
plot(GAS.res, xlab="t", ylab="Residuals", lty=1, pch=1, col="blue")
title("Residuals from the Linear Regression of Price to Production ")

par(mfrow=c(2,2))
plot(GAS.res, xlab="t", ylab="Residuals", lty=1, pch=1, col="blue"); title("Residuals of Price on Production")
plot(density(GAS.res), main="Kernel Density of Residuals")
acf(GAS.res, main="ACF of Residuals")
pacf(GAS.res, main="PACF of Residuals")

#Cross correlation Function between Price and Production
par(mfrow=c(1,1))
ccf(price.ts, production.ts, main="Cross-correlation between Price and Production")

# This shows that histogram and correlation may not be effective tools for time series data
pm = ggpairs(cbind(price.ts, production.ts))
print(pm)

#Fit an Autoregressive Model to both price and production series
GAS.ar <- ar(cbind(price.ts, production.ts), method="ols", dmean=T, intercept=F)
summary(GAS.ar) # Objects of the estimation results
GAS.ar$ar
GAS.ar$order
#==> it suggests that the best fitting VAR model is of order 26

# Diagnosis using the estimated residuals
dim(GAS.ar$res)
summary(GAS.ar$res)
GAS.ar$res #list the residual

ggpairs(GAS.ar$res) # the residuals do look "fairly" normal and
                   # not correlated with each other

ts.plot(GAS.ar$res[,1], GAS.ar$res[,2], gpars=list(xlab="Simulated Time Period", ylab="Series Values",      lty=c(1:2), pch=c(1,4), col=c("blue","black")))
title("Estimated Resduals from the VAR(26) Model")
leg.txt <- c('Price', 'Production')
legend('topleft', leg.txt, lty=1, col=c('blue', 'black'), bty='n', cex=.75)

#par(mfrow=c(2,2))
#  ts.plot(GAS.ar$res[-c(1:3),1], col="blue", main="Price Residuals from a VAR(26) Model")
#  ts.plot(GAS.ar$res[-c(1:3),2], col="black", main="Production Residuals from a VAR(26) Model")
#  acf(GAS.ar$res[-c(1:3),1], col="blue", main="ACF of Price Residuals")
#  acf(GAS.ar$res[-c(1:3),2], main="ACF of Production Residuals")

#Fit a Vector Vector AR Model
#GAS.var <- VAR(cbind(price.ts,production.ts), p=3, type="trend")
x <- cbind(price.ts,production.ts)
GAS.var <- VAR(x, p=1, type="trend")
summary(GAS.var)
coef(GAS.var)

# 24-Step ahead forecast or 1-year forecast
GAS.pred <- predict(GAS.var, n.ahead=24)
GAS.pred

price.pred <- ts(GAS.pred$fcst$price.ts[,1], st=c(2012,2,1), fr=12)
production.pred  <- ts(GAS.pred$fcst$production.ts[,1], st=c(2012,2,1), fr=12)

#Plot without CI info
ts.plot(cbind(window(price.ts, start=1978), price.pred), lty=1:2 )
ts.plot(cbind(window(production.ts, start=1978), production.pred), lty=1:2)

VARselect(x, lag.max=30, type="both")
summary(fit <- VAR(x, p=2, type="both"))

summary(VAR(x, p=1, type="both"))  # "both" fits constant + trend

#Run an Autocorrelation on the residuals
acf(resid(fit), 52)
serial.test(fit, lags.pt=12, type="PT.adjusted")

#Plot with 95% CI
fit.pr = predict(fit, n.ahead = 24, ci = 0.95)  # 2 years ahead
fanchart(fit.pr)  # plot prediction + error

```

Task 3: Compare the accuracy of the two models' forecasting results. Also, compare and contrast the results of these two models. Is one model better than the other? What metric(s) do you use to measure whether one model is "better" than the other? Why or why not? Explain the pros and cons of each of the models in this specific context.

```{r}



```

